---
title: "Class 11: Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1.1 PDB Statistics 

Download pdb stats data as CSV file from https://www.rcsb.org/stats/summary
```{r}
pdbstats <- read.csv("Data Export Summary.csv", row.names = 1)
library(knitr)
kable(pdbstats)
```

```{r}
#total number of structures
tot.st <- sum(pdbstats$Total)
#total number of structures 
percent <- round(pdbstats$Total/tot.st*100, 2)
nstats <- pdbstats
nstats$Percent <- percent
kable(nstats)
```


Question 1:
There are `r percent[1]`% X-ray structures and `r percent[2]`% NMR strutures in the PDB data base as of `r Sys.Date()`.

```{r}
percent.pro <- round(sum(pdbstats$Proteins)/tot.st*100, 2)

```

There are `r percent.pro`% protein strutures in the PDB data base as of `r Sys.Date()`.

##Section 3: Getting Started with Bio3D

Any time we want to use a package we must load it with the **library()** function. 
```{r}
library(bio3d)
```

```{r}
pdb <- read.pdb("1hsg")
pdb
head(pdb$atom)
```

```{r}
pdb$xyz
dim(pdb$xyz)
pdb$xyz[ 1, atom2xyz(1:2)]
```

Select all C-alpha atoms
```{r}
ca.inds <- atom.select(pdb, "calpha")
ca.inds

```

```{r}
#Print details of the first few selected atoms
head(pdb$atom[ca.inds$atom,])
#And selected xyz coordinates
head(pdb$xyz[,ca.inds$xyz])
```

```{r}
ligand.inds <- atom.select(pdb, "ligand")
ligand.inds
```

```{r}
pdb$atom[ligand.inds$atom,]
```

```{r}

```

##Section 4: Working with multiple PDB files


Aligning multiple structures
```{r}
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```

Extract and Align the chains we are interested in 
Print to screen a summary of the 'pdbs' object 
```{r}
pdbs <- pdbaln(files, fit = TRUE)
pdbs
```

To access the first few rows of the alignment matrix we use standard subsetting syntax for matrices in R:
```{r}
pdbs$ali[1:5, 1:8]
```

Associated residue numbers
```{r}
pdbs$resno[1:5, 1:8]
```

## PCA of the above aligned sequences

Find invariant core
```{r}
core <- core.find(pdbs)
```
Superimpose all structures to core
```{r}
pdbs$xyz = pdbfit(pdbs, core)
```
Perform PCA
```{r}
pc.xray <- pca(pdbs)
plot(pc.xray)
```
Calculate RMSD
```{r}
rd <- rmsd(pdbs)
```
## Visualize first principal component
```{r}
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

