---
title: "class09"
author: "Hannah Morris Little"
date: "10/30/2018"
output: github_document
---
Background
The goal of this hands-on session is for you to explore a complete analysis using the unsupervised learning techniques covered in the last class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass. For example radius (i.e. mean of distances from center to points on the perimeter), texture (i.e. standard deviation of gray-scale values), and smoothness (local variation in radius lengths). Summary information is also provided for each group of cells including diagnosis (i.e. benign (not cancerous) and and malignant (cancerous)).

Section 1.
Preparing the data
```{r}
url <- "https://bioboot.github.io/bimm143_W18/class-material/WisconsinCancer.csv"
wisc.df <- read.csv(url)
```

```{r}
head(wisc.df)
```
Examine your input data to ensure column names are set correctly. The id and diagnosis columns will not be used for most of the following steps. Use as.matrix() to convert the other features (i.e. columns) of the data (in columns 3 through 32) to a matrix. Store this in a variable called wisc.data.

Assign the row names of wisc.data the values currently contained in the id column of wisc.df. While not strictly required, this will help you keep track of the different observations throughout the modeling process.
```{r}
wisc.data <- as.matrix(wisc.df[,3:32])
row.names(wisc.data) <- wisc.df$id
#Take columns 3 - 32
head(wisc.data)
```


```{r}
#How would we make a vector of ones and zeroes where M = 1 and B = 0
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
table(wisc.df$diagnosis)
#Check that diagnosis vector matches M count
sum(diagnosis)
```

Exploratory Data Analysis
```{r}
#Q1. How many observations are in the dataset
nrow(wisc.df)
#Q2. How many variables are in the data set suffixed with _mean
x <- length(grep(pattern = "_mean", colnames(wisc.data)))
#Q3. How many observations are malignant. Table function will count the number of values. 
table(wisc.df$diagnosis)
```

There are `r x` mean measurements in this dataset



Section 2.
Performing PCA

It is important to check if the data need to be scaled before performing PCA. Recall two common reasons for scaling data include:

The input variables use different units of measurement.
The input variables have significantly different variances.
Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the colMeans() and apply() functions like you’ve done before.

```{r}
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
#We see that PC1 accounts for 44% of variance and PC2 accounts for 19% of variance
wisc.prF <- prcomp(wisc.data, scale = FALSE)
summary(wisc.prF)
#We see that PC1 accounts for 98% of variance
```
Q1. PC1 accounts for 44% of variance
Q2. 3 PCs are required to describe at least 70% variance in the data
Q3. 7 PCs are required to describe at least 90% variance in the data

```{r}
biplot(wisc.pr)
```

```{r}
attributes(wisc.pr)
#There is no color for values that are 0 so they are colored white. Instead just add one. Benign now = 1 and malignant = 2 so benign = black and malignant = red
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis+1, xlab = "PC1", ylab = "PC2")

```

We are seeing actual separation of malignancy and benign in the PC1 but not in the PC2

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis+1, xlab = "PC1", ylab = "PC3")
```

Same trend as seen above. Separation in PC1 but not PC2 

```{r}
plot(wisc.pr$x[,2], wisc.pr$x[,3], col = diagnosis+1, xlab = "PC2", ylab = "PC3")
```

Overall, the plots indicate that principal component 1 is capturing a separation of malignant from benign samples. This is an important and interesting result worthy of further exploration - as we will do in the next sections!

Variance explained

In this exercise, you will produce scree plots showing the proportion of variance explained as the number of principal components increases. The data from PCA must be prepared for these plots, as there is not a built-in function in base R to create them directly from the PCA model.

As you look at these plots, ask yourself if there’s an ‘elbow’ in the amount of variance explained that might lead you to pick a natural number of principal components. If an obvious elbow does not exist, as is typical in some real-world datasets, consider how else you might determine the number of principal components to retain based on the scree plot.

Calculate the variance of each principal component by squaring the sdev component of wisc.pr (i.e.  wisc.pr$sdev^2). Save the result as an object called pr.var.

Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components. Assign this to a variable called pve and create a plot of variance explained for each principal component

```{r}
pr.var <- wisc.pr$sdev^2
tot.var <- sum(pr.var)
pve <- pr.var/tot.var
pve
plot(pve, xlab = "Principle Component", ylab = "Proportion of Variance Explained", ylim = c(0,1), typ = "o")
```
```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

```

```{r}
plot(cumsum(pve), xlab = "Principle Component", 
     ylab = "Cumulative Propotion of Variance Explained", 
     ylim = c(0,1), type = "o")
```


```{r}
par(mfrow = c(1,2))
plot(cumsum(pve), xlab = "Principle Component", 
     ylab = "Cumulative Propotion of Variance Explained", 
     ylim = c(0,1), type = "o")
plot(pve, xlab = "Principle Component", ylab = "Proportion of Variance Explained", ylim = c(0,1), typ = "o")
```

Section 3.
Hierarchical clustering of case data

The goal of this section is to do hierarchical clustering of the observations. Recall from our last class that this type of clustering does not assume in advance the number of natural groups that exist in the data.

As part of the preparation for hierarchical clustering, the distance between all pairs of observations are computed. Furthermore, there are different ways to link clusters together, with single, complete, and average being the most common linkage methods.

Scale the wisc.data data and assign the result to data.scaled.




```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
```

Selecting number of clusters
In this section, you will compare the outputs from your hierarchical clustering model to the actual diagnoses. Normally when performing unsupervised learning like this, a target variable (i.e. known answer or labels) isn’t available. We do have it with this dataset, however, so it can be used to check the performance of the clustering model.

When performing supervised learning - that is, when you’re trying to predict some target variable of interest and that target variable is available in the original data - using clustering to create new features may or may not improve the performance of the final model.

This exercise will help you determine if, in this case, hierarchical clustering provides a promising new feature.

Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable  wisc.hclust.clusters.

```{r}
wisc.hclust.clusters4 <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters4, diagnosis)
```
Here we picked four clusters and see that cluster 1 largely corresponds to malignant cells (with diagnosis values of 1) whilst cluster 3 largely corresponds to benign cells (with diagnosis values of 0).

Section 4.
K-means clustering and comparing results

Section 5.
Clustering on PCA results

In this final section, you will put together several steps you used earlier and, in doing so, you will experience some of the creativity and open endedness that is typical in unsupervised learning.

Recall from earlier sections that the PCA model required significantly fewer features to describe 70%, 80% and 95% of the variability of the data. In addition to normalizing data and potentially avoiding over-fitting, PCA also uncorrelates the variables, sometimes improving the performance of other modeling techniques.

Let’s see if PCA improves or degrades the performance of hierarchical clustering.

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with complete linkage. Assign the results to wisc.pr.hclust.


```{r}
d.pr <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(d.pr)
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 4)
table(wisc.pr.hclust.clusters, diagnosis)
table(wisc.hclust.clusters4, diagnosis)
```

Section 6.
Sensitivity refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples.

Specificity relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign.

Q16. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?


##Bonus Section
Predicting with our PCA model 
Take new patient data and apply our PCA model 

```{r}
## Predicting Malignancy Of New samples
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata = new)
plot(wisc.pr$x[,1:2], col = diagnosis + 1)
points(npc[,1], npc[,2], col = "blue", pch = 16, cex = 2)
```

One patient we should worry more about than the other. Patient 2 we should worry more about. 


