---
title: "class 08"
author: "Hannah Morris Little"
date: "10/25/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

Or first example with **kmeans()** function.

```{r}
# Example plot to see how Rmarkdown workds
plot(1:10, typ = "l")
```

Back to kmeans...
Option, command, i <- puts in R chunk

```{r}
# Generate some example data for clustering
# 30 numbers centered around -3 and 30 numbers centered aroung 3
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x = tmp, y = rev(tmp))
plot(x)
```
 Use the kmeans() function setting k to 2 and nstart=20
 Inspect/print the results
 
 Q. How many points are in each cluster?
 Q. What ‘component’ of your result object details
 Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

```{r}
# cluster size?
k$size
table(k$cluster)
```

```{r}
# cluster center?
k$centers
```

```{r}
# cluster assignment/membership? 
k$cluster
```

```{r}
# Color by cluster in your palette of choice and display cluster centers in green 
plot(x, col = k$cluster)
points(k$centers, col = "green", pch = 20, cex = 3)
```

 Q. Repeat for k=3, which has the lower tot.withinss? 
 
```{r}
k3 <- kmeans(x, centers = 3, nstart = 20)
k4 <- kmeans(x, centers = 4, nstart = 20)
k5 <- kmeans(x, centers = 5, nstart = 20)
k6 <- kmeans(x, centers = 6, nstart = 20)

k6$tot.withinss
k5$tot.withinss
k4$tot.withinss
k3$tot.withinss
k$tot.withinss

plot(c (k6$tot.withinss, k5$tot.withinss, k4$tot.withinss, k3$tot.withinss, k$tot.withinss))
```

## Hierarchical clustering
```{r}
dist_matrix <- dist(x)
hc <- hclust(d = dist_matrix)
hc
plot(hc)
abline(h = 8, col = "red")
cutree( hc, k = 2)
```

```{r}
x <- rbind( 
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
  rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
x
k1 <- kmeans(x, centers = 3)
k1$centers
palette(c("blue", "red", "green"))
col <- as.factor( rep(c("c1", "c2", "c3"), each = 50))
plot(x, col = col)
points(k1$centers, col = "black", pch = 20, cex = 3)
```

Q. Use the dist(), hclust(), plot(), and cutree()
functions to return 2 and 3 clusters
```{r}
dist_matrix1 <- dist(x)
hc1 <- hclust( d = dist_matrix1)
hc1
plot(hc1)
abline( h = 2, col = "red")
grp.3 <- cutree( hc1, k = 3)
grp.2 <- cutree(hc1, k = 2)

table(grp.2)
table(grp.3, col)

plot(x, col = grp.3)
plot(x, col = grp.2)
```

## PCA
```{r}
#initialize a blank 100 row by 10 column matrix
mydata <- matrix( nrow = 100, ncol = 10)
#label the rows gene1 : gene100
rownames(mydata) <- paste("gene", 1:100, sep = "")
#label the first 5 columns WT1:WT5 and the last 5 KO1:KO5
colnames(mydata) <- c( paste("wt", 1:5, sep = ""),
 paste( "ko", 1:5, sep = "") )
#Fill in some fake read counts
for(i in 1:nrow( mydata)) {
 wt.values <- rpois(5, lambda = sample( x = 10:1000, size = 1))
 ko.values <- rpois(5, lambda = sample(x = 10:1000, size = 1))
 mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
head(t(mydata))

#Lets do PCA
pca <- prcomp(t(mydata), scale = TRUE)
#What are the attributes of prcomp()
attributes(pca)

#A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
#To get the variance you need to square the sdev
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
#Kind of like the scree plot, we see that the drop is after pc 1
pca.var.per
barplot(pca.var.per, main = "Scree Plot", xlab = "Principle Component", ylab = "Percent Variation")

```
```{r}
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot( pca$x[,1], pca$x[,2], col = colvec, pch = 16,
 xlab = paste0("PC1 (", pca.var.per[1], "%)"),
 ylab = paste0("PC2 (", pca.var.per[2], "%)")) 
```
```{r}
loading_scores <- pca$rotation[,1]

gene_scores <- abs(loading_scores)
gene_score_ranked <- sort(gene_scores, decreasing = TRUE)
top_5_genes <- names(gene_score_ranked[1:5])
pca$rotation[top_5_genes,1]
```

```{r}
x <- read.csv("UK_foods.csv")

rownames(x) <- x[,1]
x <- x[,-1]
knitr::kable(x, caption = "The full UK foods data table")

dim(x)

par(mar = c(20, 4, 4, 2))
heatmap(as.matrix(x))
```



