---
title: "Class 13: Structural Bioinformatics II"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get HIV-Pr structure from PDB database

We will work with the struture 1HSG

```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
```

Read this into R and examine composition

```{r}
hiv <- read.pdb(file.name)
hiv
```

Question 1: The ligand is MK1 and the second nonprotein residue is HOH

## Prepare initial protein and ligand input files

Split into protein only and a ligand only

We will use the `trim.pdb()` function to split our input structure

```{r}
prot <- trim.pdb(hiv, "protein")
prot
```

Write out a new protein-only pdb file

```{r}
write.pdb(prot, file = "1hsg_protein.pdb")
```

Do the same for ligand-only

```{r}
lig <- trim.pdb(hiv, "ligand")
lig
write.pdb(lig, file = "ligand.pdb")
```

## Using AutoDockTools to setup protein docking input

## Prepare docking configuration file

## Docking with Vina
We run this command:
`~/Downloads/autodock_vina_1_1_2_mac/bin/vina --config config.txt --log log.txt`

##Read docking results

We got a file all.pdbqt that we need to make into a PDB format.

```{r}
res <- read.pdb("all.pdbqt", multi = TRUE)
write.pdb(res, "results.pdb")
res
```

Qualitatively the crystal binding mode is nearly reproduced by the best docking conformation accouding to AutoDock Vina.

But what about qunatitatively? We can use `rmsd()` to quantify the comparison. 

```{r}
ori <- read.pdb("ligand.pdbqt")
rmsd(ori, res)
```

